\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-06 paper/ms.tex
a change
Further edits to manuscript.
intro edits

\dfmtodo{Talk about LSST here too!}

In the astrophysical literature, Gaussian Processes
\citep[GPs;][]{Rasmussen:2006} have been used to model stochastic variability
in light curves of stars \citep{Brewer:2009}, active galactic nuclei
\citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987, Bond:1999, Wandelt:2003}, correlated instrumental noise
\citep{Gibson:2012}, and spectroscopic calibration \citep{Czekala:2017,
Evans:2015}.
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the cube of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
$\sim10^{3-5}$ observations each~--~this scaling is prohibitive.

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
\dfmtodo{This method is a generalization of a method developed by
\citet{Ambikasaran:2015}
 ({\bf This is no longer correct -different Ambikasaran result!})}
that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\dfmtodo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
system that can be solved efficiently using a sparse or banded solver. Replace
with(?): The matrix can be expressed as the sum of semi-separable matrices, each
of which is constructed from the outer product of one-dimensional vectors.  These
are a type of low-rank matrix, which may be solved by Cholesky decomposition
using a novel technique developed by one of us (SA).}
As we will discuss in the following pages, this choice of kernel function is
not as restrictive as it first seems because it arises naturally in some
physical systems and it can be used as an effective model in many other cases.
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

Our main expertise lies in the field of exoplanet discovery and
characterization where GPs have become a model of choice.
We are confident that the method described in the following pages will benefit
this field but we also expect that there will be applications in other
branches of astrophysics and beyond.
In \sect{examples-real}, we present applications of the method to research
problems in stellar rotation, asteroseismic analysis, and exoplanet transit
fitting.
Some readers might consider starting with that section for some motivating
examples before delving into the detailed derivations of the earlier sections.

In the following pages, we motivate the general problem of GP modeling,
describe the \dfmtodo{(delete) previously published} scalable method
\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
the model's application on various real and simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo}
\citep{Foreman-Mackey:2017}.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the cost of
computing the inverse and determinant of the matrix $K_\bvec{\alpha}$ is
$\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
This means that for large datasets, every evaluation of the likelihood
quickly becomes computationally intractable.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

In the following section, we present a method for substantially improving this
scaling in many circumstances.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.
\response{Throughout the paper, when referring to the \celeriteterm\ model, we
typeset \celeriteterm\ in italics and, in reference to the implementation of
this model, we typeset \celerite\ in sans-serif.}}
The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.
We use $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the
\celerite\ method can be applied to \emph{any} one-dimensional dataset.
Furthermore, the covariance function for the \celerite\ method is
``stationary''.
In other words, $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of $\tau_{nm}
\equiv |t_n - t_m|$.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(N\,J^2)$, where $J$ is the number of components
in the mixture.

@@

\section{Examples}

@@

\subsection{Example 1: Recovery of a celerite process}

@@

\subsection{Example 2: Inferences with the ``wrong'' model}

@@

\subsection{Example 4: Asteroseismic oscillations}

@@

\subsection{Example 5: Exoplanet transit fitting}

@@

\section{Comparisons to other methods}

>>>>

In the astrophysical literature, Gaussian Processes
\citep[GPs;][]{Rasmussen:2006} have been used to model stochastic variability
in light curves of stars \citep{Brewer:2009}, active galactic nuclei
\citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987, Bond:1999, Wandelt:2003}, correlated instrumental noise
\citep{Gibson:2012}, and spectroscopic calibration \citep{Czekala:2017,
Evans:2015}.
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
The cost of computing a general GP likelihood scales as the cube of the number
of data points $\mathcal{O}(N^3)$ and in the current era of large time domain
surveys~--~with as many as $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~--~this scaling is prohibitive.

\response{Existing astronomical time series datasets have already reached the
limit where na\"ive application GP models is no longer tractable.
NASA's \kepler\ Mission \citep{Borucki:2010}, for example, measured light
curves with more than 60,000 observations each for about 190,000 stars.
Current and forthcoming surveys such as \project{K2} \citep{Howell:2014},
\tess\ \citep{Ricker:2014}, \lsst\ \citep{Ivezic:2008}, \project{WFIRST}
\citep{Spergel:2015}, and PLATO \citep{Rauer:2014} will continue to produce
similar or larger data volumes.}

In this paper, we present a method for directly and exactly computing a class
of GP models that scales linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
\response{Unlike earlier linear methods using Kalman filters \citep[for
example,][]{Kelly:2014}, this novel algorithm exploits the semi-separable
structure of a specific class of covariance matrices to directly factorize and
solve the system.
}
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, the measurements don't need to be evenly spaced and the uncertainties
can be heteroscedastic.

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semi-separable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in some physical systems and we demonstrate it can be used as
an effective model in some other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

Our main expertise lies in the field of exoplanet discovery and
characterization where GPs have become a model of choice.
We are confident that the method described in the following pages will benefit
this field, but we also expect that there will be applications in other
branches of astrophysics and beyond.
In \sect{examples}, we present applications of the method to research problems
in stellar rotation (\sect{rotation}), asteroseismic analysis (\sect{astero}),
and exoplanet transit fitting (\sect{transit}).
Some readers might consider starting with these section for motivating
examples before delving into the detailed derivations of the earlier sections.

In the following pages, we motivate the general problem of GP modeling, derive
our novel direct solver, and demonstrate the model's application on real and
simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo}
\citep{Foreman-Mackey:2017}.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
A point estimate for the values of the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
\response{Furthermore}, probabilistic constraints on $\bvec{\theta}$ and
$\bvec{\alpha}$ can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the cost of
computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
$\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
Furthermore, the storage requirements also scale as $\mathcal{O}(N^2)$.
This means that for large datasets, every evaluation of the likelihood quickly
becomes computationally intractable, and representing the matrix in memory can
become expensive.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

\response{This is not a purely academic concern.
While the cost of directly evaluating the GP likelihood using a tuned linear
algebra library\footnote{For comparisons throughout this paper, we use the
Intel Math Kernel Library (\url{https://software.intel.com/en-us/intel-mkl})
running on a recent Apple MacBook Pro.} with a dataset of $N=1024$
measurements is less than a tenth of a second, the same calculation with
$N=8192$ takes over 8~seconds.
Furthermore, most optimization and sampling routines are iterative, requiring
many evaluations of this model to perform parameter estimation or inference.
Existing datasets from astronomical surveys like \kepler\ and \project{K2}
include tens of thousands of observations for hundreds of thousands of
targets, and upcoming surveys like \lsst\ are expected to produce thousands or
tens of thousands of measurements each for billions of astronomical sources
\citep{Ivezic:2008}.
Scalable methods will be required if we ever hope to use GP models with these
datasets.
}

In the following section, we present a method for improving the cubic scaling
for some use cases.
We call our method and its implementations \celerite.\footnote{The name
\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
of light in a vacuum.
\response{Throughout the paper, when referring to the \celeriteterm\ model, we
typeset \celeriteterm\ in italics and, in reference to the implementation of
this model, we typeset \celerite\ in sans-serif.}}
The \celerite\ method requires using a specific model
for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
has some limitations, we demonstrate in subsequent sections that it can
increase the computational efficiency of many astronomical data analysis
problems.
The main limitation of this method is that it can only be applied to
one-dimensional datasets, where by ``one-dimensional'' we mean that the
 \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
t_n$.
We use $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the
\celerite\ method can be applied to \emph{any} one-dimensional dataset.
Furthermore, the covariance function for the \celerite\ method is
``stationary''.
In other words, $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of $\tau_{nm}
\equiv |t_n - t_m|$.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
\end{eqnarray}
\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
kernel is semi-separable.}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture.
\response{In the following section, we derive a novel method with the same
$\mathcal{O}(N\,J^2)$ scaling that can be used when
$k_\bvec{\alpha}(\tau_{nm})$ is positive semi-definite.}

@@

\section{Examples}\sectlabel{examples}

@@

\subsection{Example 1: Recovery of a celerite process}\sectlabel{example1}

@@

\subsection{Example 2: Inferences with the ``wrong'' model}\sectlabel{example2}

@@

\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}

@@

\subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}

@@

\section{Comparisons to other methods}\sectlabel{compare}
